---
title: "02-Time-Series-Decomposition"
author: "Hector Gavilanes"
format: html
editor: visual
self-contained: true
execute: 
  warning: false
  message: false
---

Plot the data. Inspection of the graph may suggest the possibility of representing the data as follow:

## The classical decomposition

$$
X_t = m_t + Y_t
$$

where

-   $m_t$ is the trend component

-   $s_t$ is the seasonal component

-   $Y_t$ random noise component / Residuals

if seasonality and noise fluctuations (variability) appear to increase with the level of the process (trend), then use a preliminary transformation of the data (natural logarithm,...).

We are going to learn how to estimate and eliminate the trend and the seasonal components in the hope that the residual $Y_t$ will be a **stationary time series.**

```{r setup, echo=FALSE}
# install packages if needed
#install.packages("itsmr")
```

```{r, warnings = FALSE, message = FALSE}
# Load packages
library(itsmr)
library(tidyverse)
library(tsibble)
library(feasts)
```

# Trend Estimation

```{r}
strikes = read.table("http://users.stat.umn.edu/~kb/classes/5932/data/strikes.txt", header = FALSE)

# get data to tsibble format
strikes = strikes %>% 
  mutate(Year=seq(1951,1980,1)) %>%
  as_tsibble(index = Year) 

# plot the time series
strikes %>%
  autoplot(V1) + 
  labs(y="Number of Strikes (thousands)",
       title="Strikes in USA in 1951-1980")
```

```{r}
# plot the ACF
strikes %>%
  ACF(V1, lag_max = 20) %>%  
  autoplot()
```

# Trend Estimation

## Smoothing Moving Average

```{r}
strikes %>%
  mutate(ma.trend3 = smooth.ma(V1, 3),
         ma.trend5 = smooth.ma(V1, 5),
         ma.trend7 = smooth.ma(V1, 7),
         ma.trend9 = smooth.ma(V1, 9)) %>% # q=3 means averaging 2q+1 values
  pivot_longer(cols =-Year,
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)
```

# Trend Estimation

## Exponential Smoothing

```{r}
strikes %>%
  mutate(exp.trend1 = smooth.exp(V1, 0.1),# alpha=0.1 mis the smootheness parameter
         exp.trend2 = smooth.exp(V1, 0.5),
         exp.trend3 = smooth.exp(V1, 0.7),
         exp.trend4 = smooth.exp(V1, 0.9)) %>% 
  pivot_longer(cols =-Year,
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)
```

# Trend Estimation

## FFT

```{r}
strikes %>%
  mutate(fft.trend1 = smooth.fft(V1, 0.1),# 0.1 is cutoff frequency - 10% of the lowest spectrum passes
         fft.trend2 = smooth.fft(V1, 0.25)) %>% 
  pivot_longer(cols =-Year,
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)
```

# Trend Estimation 

## Regression

```{r}
strikes %>%
  mutate(reg.trend1 = trend(V1, 1),# 1 means polynmial of degree 1 
         reg.trend2 = trend(V1, 2),
         reg.trend3 = trend(V1, 3)) %>% 
  pivot_longer(cols =-Year,
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)
```

# Seasonality Estimation

```{r}
plot.ts(deaths)
```

```{r}
acf(deaths,lag.max=40)
```

```{r}
estimate.season = season(deaths, d=12)
plotc(deaths,estimate.season)
```

```{r}
r = deaths - estimate.season
test(r)
```

# Test IID hypothesis for the Residuals

Use the Resid() function to find the residuals.

```{r}
# deaths
data.model = c("season",12,"trend",1) # from examining the series
R= Resid(deaths,data.model)
# Test IDD
test(R) # need to tests to fail to reject the null hypothesis
```

```{r}
# strikes
data.model.s = c("trend",3)
R.s= Resid(strikes$V1,data.model.s)
# Test IDD
test(R.s)
```

# **Trend and Seasonality Elimination using the Differencing operator**

## **Johnson & Johnson quarterly earnings data**

```{r}
plot.ts(JohnsonJohnson)
```

```{r}
acf(JohnsonJohnson,lag.max = 50)
```

```{r, warnings = FALSE, message = FALSE}
# Transformation
library(car)
L = powerTransform(JohnsonJohnson)
Tjj = log(JohnsonJohnson)

plot.ts(Tjj)
```

```{r}
acf(Tjj,lag.max = 50)
```

There is a trend with the assumption to be quadratic. There is change in variance.

The data model:

$$
X_t = m_t + Y_t
$$

```{r}
# Estimations
jj.data.model = c("log","trend",1)
r.jj = Resid(JohnsonJohnson,jj.data.model)
test(r.jj)
```

It seems there is a season of period 4. Then:

$$
X_t = m_t + S_t + Y_t
$$

```{r}
# Estimation 
jj.data.model = c("log","trend",2,"season",4)
r.jj = Resid(JohnsonJohnson,jj.data.model)
test(r.jj)
```

By elimination we have the following:

$$
y_t = \nabla^2 \nabla_4log(X_t) 
$$

The rewrite form:

$$
y_t = (1-B)^2 (1-B^4)X_t 
$$

Elimination of season

```{r}
deseason.data = diff(log(JohnsonJohnson),lag = 4)
plot(deseason.data)
```

Trend Elimination

```{r}
residual.data = diff(deseason.data,lag = 1,differences = 2)
```

Test IID for Residuals

```{r}
test(residual.data)
```

Since at least one test rejects the hypothesis of IID, thus further analysis needs to be performed.
