---
title: "STA6856 Time Series"
subtitle: "SARIMA modeling"
author: "Hector Gavilanes"
format: html
self-contained: true
execute: 
  warning: false
  message: false
editor: visual
---

```{r}
# libraries
library(itsmr)
library(car)      #power transformation
library(aTSA)
library(forecast)
```

Fit SARIMA(p,d,q)x(P,D,Q) models to the following time series data:

1.  `airpass`

2.  `wine`

Post your work (HTML) on Canvas discussion board.

$$
\phi(B)\Phi(B^s)(1-B^s)^D(1-B)^d X_t = \theta(B)\Theta(B^s) Z_t, \quad Z_t \sim WN(0,\sigma^2)
$$

## Airpass

```{r}
# plot
plot.ts(airpass)
plotc(airpass)
```

```{r}
# ACF, PACF
acf(airpass, lag.max = 50)
pacf(airpass, lag.max = 40)
plota(airpass)
```

## Power Transformation

```{r}
# Variance change transformation
lambda = powerTransform(airpass)
lambda$roundlam
```

Lambda $\lambda$=0; thus, we can apply Log Power Transformation.

## Differencing Operator

By Elimination, we have the following:

$$
y_t = \nabla^1 \nabla_{12} +Y_t
$$

```{r}
# Eliminate components using Differencing Operator
air_diff = c("log", "diff", 12, "diff", 1)
r_diff = Resid(airpass, air_diff)
test(r_diff)
```

Reject the IID null hypothesis.

## Test Stationary

```{r}
adf.test(r_diff)
```

## SARIMA Model

### Based on the PACF

p = 0,1,2,9,10,11,13

P = 0,1,2

### Based on the ACF

q = 0:40

Q = 0,1,2,3

### Eliminate Trend and Season

d= 1

D = 1

```{r}
# SARIMA (1,1,1)x(1,1,1)12
fit1 = arima(airpass,
             order = c(1,1,1),
             seasonal = list(order=c(1,1,1), period=12))
test(fit1$residuals)
confint(fit1)
```

### Auto-fit

```{r}
# Auto fit
bestmodel = auto.arima(airpass, trace = T, seasonal = T)
```

```{r}
test(bestmodel$residuals)
```

Using the best model (4,1,2) to remove seasonality for stationary models.

```{r}
# SARIMA (4,1,2)x(1,1,1)12
fit2 = arima(airpass,
             order = c(4,1,2),
             seasonal = list(order=c(1,1,1), period=12))
test(fit2$residuals)
confint(fit2)
```

Failed to reject the null hypothesis. The model is stationary.

```{r}
# SARIMA (2,1,2)x(1,1,0)12
fit3 = arima(airpass,
             order = c(2,1,2),
             seasonal = list(order=c(1,1,0), period=12))
test(fit3$residuals)
confint(fit3)
```

FTR the null hypothesis. The model is not stationary.

```{r}
# SARIMA (4,1,2)x(1,1,2)12
fit4 = arima(airpass,
             order = c(4,1,2),
             seasonal = list(order=c(1,1,2), period=12))
test(fit4$residuals)
confint(fit4)
```

FTR the null hypothesis. The model is stationary.

```{r}
# SARIMA (5,1,2)x(1,1,1)12
fit5 = arima(airpass,
             order = c(5,1,2),
             seasonal = list(order=c(1,1,1), period=12))
test(fit5$residuals)
confint(fit5)
```

```{r}
# SARIMA (13,1,2)x(1,1,0)12
fit6 = arima(airpass,
             order = c(13,1,2),
             seasonal = list(order=c(1,1,0), period=12))
test(fit6$residuals)
confint(fit6)
```

FTR the null hypothesis. The model is stationary.

## SARIMA Comparison

```{r}
# Comparison
AIC.values = c(
  autoModel = bestmodel$aic,
  fit1 = fit1$aic,
  fit2 = fit2$aic,
  fit3 = fit3$aic,
  fit4 = fit4$aic,
  fit5 = fit5$aic,
  fit6 = fit6$aic
)
AIC.values
```

## SARIMA Equation

The equation of SARIMA (13,1,2)x(1,1,0)12 is:

```{r}
round(fit2$coef, digits = 3)
```

$$
(1-0.357B+0.817B^2+0.229^3-0.110B^4 -0.869^{12})(1-B^{12})(1-B)X_t
$$

$$
(1+0.026B-0.974B^2)(1-0.771^{12})Z_t
$$

$$
Z_t \sim WN(0,116.9)
$$

The equation of SARIMA (4,1,2)x(1,1,0)12

```{r}
round(fit6$coef, digits = 3)
```

$$
(1+1.262B -0.172B^2 -0.219^3 -0.080B^4 +0.322B^5 -0.184B^6 -0.165B^7 +0.120B^8 +0.268B^9 -0.499B^{10} + 0.407B^{11} +0.226B^{12} - 0.411B^{13} -0.609^{12}) (1-B^{12}) (1-B) X_t
$$

$$
(1-1.679B + 0.700B^2) Z_t
$$

$$
Z_t \sim WN(0,97.34)
$$

# Wine

```{r}
plotc(wine)
plot.ts(wine)
```

```{r}
# ACF, PACF
acf(wine, lag.max = 30)
pacf(wine, lag.max = 30)
plota(wine)
```

## Comments on Trend, Seasonality, Variance

The Wine time series has upward trend, seasonality (d=12) based on ACF, and variance change.

The Wine model:

$$
X_t = m_t + s_t + Y_t
$$

## Power Transformation

```{r}
# Variance change transformation
lambda = powerTransform(wine)
lambda$roundlam
```

Lambda $\lambda$ = 0.5; thus, square root transformation needed.

```{r}
wine <- basicPower(wine, lambda$roundlam)
```

```{r}
# Seasonality
wine_season = season(wine, d=12)
plotc(wine, wine_season)
```

## Differencing Operator

By Elimination, we have the following:

$$
y_t = \nabla_{12} +Y_t
$$

```{r}
# Eliminate components using Differencing Operator
wine_diff = c("diff", 12, "diff", 1)
wine_r_diff = Resid(wine, wine_diff)
test(wine_r_diff)
```

Reject the IID null hypothesis.

## Test Stationary

```{r}
adf.test(wine_r_diff)
```

## SARIMA Model

### Based on the PACF

p = 0,1,7,8,9,11,12,13,17

P = 0,1,2

### Based on the ACF

q = 0:5, 7:16, 20:28

Q = 0,1,2

### Eliminate Trend and Season

d= 1

D = 1

```{r}
# SARIMA (1,1,1)x(1,1,1)12
wine_fit1 = arima(wine,
             order = c(1,1,1),
             seasonal = list(order=c(1,1,1), period=12))
test(wine_fit1$residuals)
confint(wine_fit1)
```

FTR the null hypothesis. The model is stationary.

### Auto-Fit

```{r}
# Auto fit
wine_bestmodel = auto.arima(wine, trace = T, seasonal = T)
```

```{r}
test(wine_bestmodel$residuals)
```

Using the best model (1,1,1) to remove seasonality for stationary models.

```{r}
# SARIMA (1,1,1)x(1,1,2)12
wine_fit2 = arima(wine,
             order = c(1,1,1),
             seasonal = list(order=c(1,1,2), period=12))
test(wine_fit2$residuals)
confint(wine_fit2)
```

FTR the null hypothesis. The time series model is stationary.

```{r}
# SARIMA (1,1,1)x(0,1,1)12
wine_fit3 = arima(wine,
             order = c(1,1,1),
             seasonal = list(order=c(0,1,1), period=12))
test(wine_fit3$residuals)
confint(wine_fit3)
```

FTR the null hypothesis. The time series model is stationary.

```{r}
# SARIMA (1,1,1)x(1,1,0)12
wine_fit4 = arima(wine,
             order = c(1,1,1),
             seasonal = list(order=c(1,1,0), period=12))
test(wine_fit4$residuals)
confint(wine_fit4)
```

FTR the null hypothesis. The time series model is stationary.

```{r}
# SARIMA (1,1,1)x(0,1,2)12
wine_fit5 = arima(wine,
             order = c(1,1,1),
             seasonal = list(order=c(0,1,2), period=12))
test(wine_fit5$residuals)
confint(wine_fit5)
```

FTR the null hypothesis. The time series model is stationary.

## SARIMA Comparison

```{r}
# Comparison
AIC.values = c(
  autoModel = bestmodel$aic,
  wine_fit1 = wine_fit1$aic,
  wine_fit2 = wine_fit2$aic,
  wine_fit3 = wine_fit3$aic,
  wine_fit4 = wine_fit4$aic,
  wine_fit5 = wine_fit5$aic
)
AIC.values
```

## SARIMA Equation

```{r}
round(coefficients(wine_fit3), digits = 3)
```

$$
(1+0.094B)(1-B^{12})(1-B)X_t
$$

$$
(1-0.847B)(1-0.681B^{12})Z_t
$$

$$
Z_t \sim WN(0,5.079)
$$

Another way to write the equation:

$$
\phi(B)\Phi(B^{12}) (1-B^{12}) (1-B) X_t = \theta(B)\Theta(B^{12}) Z_t \quad Z_t \sim WN(0, \sigma^2)
$$

$$
\phi=0.094, \quad \theta=-0.847, \quad \Phi=0, \quad \Theta = -0.681
$$

$$
Z_t \sim WN(0,5.079)
$$
