---
title: "STA6856 Time Series"
subtitle: "Assignment 3"
author: "Hector Gavilanes"
format: html
embed-resources: true
execute: 
  warning: false
  message: false
editor: visual
---

We are going to fit an SARIMA(p, d, q) × (P, D, Q)s model to a dataset. We will be using the Monthly Live birth in thousands in the U.S., 1948 to 1979.

This data is given by birth in package `astsa`. You can proceed as follows:

```{r, message=FALSE, warning=FALSE}
library(itsmr)
library(astsa)
library(aTSA)     #adf.test, forecast()
library(forecast)
library(fpp3)
library(car)      #power transformation
```

## 1. Examine the plots of the time series, its ACF, and its PACF.

```{r}
plot(birth)
# ACF, PACF
acf(birth, lag.max = 80)
pacf(birth, lag.max = 50)
plota(birth)
```

-   There is an an upward trend at the beggining, but it has a correction (downward trend) after 1960.

-   There is seasonality of 12.

-   It seems that there is no drastic variance changes.

## 2. Do a transformation and eliminate the trend and/or seasonal components, if applicable.

```{r}
# Variance change transformation
lambda = powerTransform(birth)
lambda$roundlam

# Seasonality
birth_season = season(birth, d=12)
plotc(birth_season)
```

Lambda is 1, then $\lambda$ = y. No transformation required. However, there is seasonality, d = 12.

## 3. Test IID hypothesis for the residuals R1.

```{r}
# Differencing Operator
birth_diff = c("diff", 12, "diff", 1)

# IID hypothesis test
birth_r_diff = Resid(birth, birth_diff)
test(birth_r_diff)
```

-   Reject the IID null hypothesis. The time series is not stationary.

## 4. If the IID hypothesis is rejected then test the stationary hypothesis.

```{r}
# Stationary Test
adf.test(birth_r_diff)
```

-   FTR the null hypothesis. The time series is stationary.

-   No differencing is needed.

## 5. If the stationary hypothesis is not rejected then based on ACF and PACF of the residuals list the potential models.

### Based on the PACF

P = 0,1,2,5,7:14,16,24,25

p = 0,1,2,3

### Based on the ACF

q = 0,1:72

Q = 0,1:6

### Eliminate Trend & Season

d = 1 (differencing non-seasonal)

D = 1 (differencing seasonal)

## 6. Fit and compare the models using AIC Present the result in a table as follows (give at least 5 models), including the auto.arima model.

-   

    | Models  | p   | d   | q   | P   | D   | Q   | $\sigma^2$ | AIC     | IDD hypothesis |
    |---------|-----|-----|-----|-----|-----|-----|------------|---------|----------------|
    | Autofit | 0   | 1   | 2   | 1   | 1   | 1   | 46.1       | 2419.86 | FTR            |
    | 1       | 7   | 1   | 1   | 2   | 1   | 1   | 44.54      | 2424.08 | FTR            |
    | 2       | 14  | 1   | 1   | 2   | 1   | 1   | 43.07      | 2431.21 | FTR            |
    | 3       | 5   | 1   | 1   | 2   | 1   | 1   | 41.91      | 2441.46 | FTR            |
    | 4       | 5   | 1   | 5   | 1   | 1   | 2   | 43.99      | 2427.1  | FTR            |
    | 5       | 0   | 1   | 2   | 1   | 1   | 2   | 43.89      | 2420.43 | FTR            |

### Autofit

```{r}
# Auto fit
birth_bestmodel = auto.arima(birth, trace = T)
# SARIMA (0,1,2)x(1,1,1)12
```

### Fit 1

```{r}
# SARIMA (7,1,1)x(1,1,1)12
birth_fit1 = arima(birth,
             order = c(7,1,1),
             seasonal = list(order=c(1,1,1), period=12))

```

### Fit 2

```{r}
# SARIMA (14,1,1)x(2,1,1)12
birth_fit2 = arima(birth,
             order = c(14,1,1),
             seasonal = list(order=c(2,1,1), period=12))

```

### Fit 3

```{r}
# SARIMA (5,1,1)x(2,1,1)12
birth_fit3 = arima(birth,
             order = c(5,1,1),
             seasonal = list(order=c(2,1,1), period=12))

```

### Fit 4

```{r}
# SARIMA (5,1,5)x(1,1,2)12
birth_fit4 = arima(birth,
             order = c(5,1,5),
             seasonal = list(order=c(1,1,2), period=12))

```

### Fit 5 - Best Model

```{r}
# SARIMA (0,1,2)x(1,1,2)12
birth_fit5 = arima(birth,
             order = c(0,1,2),
             seasonal = list(order=c(1,1,2), period=12))

```

## 7. Diagnostic check of the residuals from the SARIMA models being fitted.

### Autofit

```{r}
# autofit
test(birth_bestmodel$residuals)
confint(birth_bestmodel)
```

### Fit 1

```{r}
# SARIMA (7,1,1)x(1,1,1)12
test(birth_fit1$residuals)
confint(birth_fit1)
```

### Fit 2

```{r}
# SARIMA (14,1,1)x(2,1,1)12
test(birth_fit2$residuals)
confint(birth_fit2)
```

### Fit 3

```{r}
# SARIMA (5,1,1)x(2,1,1)12
test(birth_fit3$residuals)
confint(birth_fit3)
```

### Fit 4

```{r}
# SARIMA (5,1,5)x(1,1,2)12
test(birth_fit4$residuals)
confint(birth_fit4)
```

### Fit 5

```{r}
# SARIMA (0,1,2)x(1,1,2)12
test(birth_fit5$residuals)
confint(birth_fit5)
```

## 8. Select a final SARIMA model and type its equation.

```{r}
# Comparison
AIC.values = c(
  automodel = birth_bestmodel$aic,
  fit1 = birth_fit1$aic,
  fit2 = birth_fit2$aic,
  fit3 = birth_fit3$aic,
  fit4 = birth_fit4$aic,
  fit5 = birth_fit5$aic
)
AIC.values
```

-   The best model is the auto-fit model.

-   SARIMA (0,1,2)x(1,1,1)12

    $$
    (B)(1-0.1018B^{12})(1-B)X_t= (1 + 0.3984B + 0.1632B^2)(1+0.8434^{12})(1-B)Z_t \quad Z_t \sim WN(0, 46.1) 
    $$

## 9. Split data to two data sets: training (367 obs.) and testing (the most recent/last 6 obs.)

```{r}
# Training Split
birth_training = birth[1:367]

# Creating a time series
birth_training_ts = ts(birth_training,start=1,frequency = 12)
```

## 10. Fit a SARIMA model to the training series and forecast the next 6 observations.

```{r}
# SARIMA (1,1,1)x(1,1,1)12
fit1_forecast = arima(birth_training,
            order = c(7,1,1),
            seasonal = list(order=c(1,1,1),period=12))

# SARIMA (1,1,1)x(2,1,1)12
fit2_forecast  = arima(birth_training,
            order = c(14,1,1),
            seasonal = list(order=c(2,1,1),period=12))

# SARIMA (1,1,1)x(2,1,1)12
fit3_forecast  = arima(birth_training,
            order = c(24,1,1),
            seasonal = list(order=c(2,1,1),period=12))

# SARIMA (1,1,1)x(2,1,1)12
fit4_forecast  = arima(birth_training,
            order = c(5,1,5),
            seasonal = list(order=c(1,1,2),period=12))

# SARIMA (0,1,1)x(1,1,2)12
fit5_forecast  = arima(birth_training,
            order = c(0,1,2),
            seasonal = list(order=c(1,1,2),period=12))

# Auto fit
bestmodel_forecast  = auto.arima(birth_training, trace= T, seasonal = T)
```

```{r}
# Best Model
bestmodel_forecast = arima(birth_training, order = c(1,1,0))
# Autofit
f.auto = aTSA::forecast(object = bestmodel_forecast,lead = 6, output = FALSE)

```

```{r}
f1 = aTSA::forecast(object = fit1_forecast,lead = 6,output = FALSE)
f2  = aTSA::forecast(object = fit2_forecast,lead = 6,output = FALSE)
f3  = aTSA::forecast(object = fit3_forecast,lead = 6,output = FALSE)
f4  = aTSA::forecast(object = fit4_forecast,lead = 6,output = FALSE)
f5  = aTSA::forecast(object = fit5_forecast,lead = 6,output = FALSE)

```

## 11. Run ARAR and Holt-Winters algorithms to predict the same number of observations.

```{r}
# ARAR
arar.prd = arar(birth_training, h=6)
```

```{r}
# Holt-Winters
HWmodel = HoltWinters(birth_training_ts,seasonal = "mult")
HW.prd =forecast::forecast(HWmodel,6)
```

## 12. Compare the models in terms of MAPE.

```{r, warning=FALSE}
library(Metrics)
MAPE1 = mape(birth[368:373],f1[,2])*100
MAPE2 = mape(birth[368:373],f2[,2])*100
MAPE3 = mape(birth[368:373],f3[,2])*100
MAPE4 = mape(birth[368:373],f4[,2])*100
MAPE5 = mape(birth[368:373],f5[,2])*100
MAPEauto = mape(birth[368:373],f.auto[,2])*100
MAPEarar = mape(birth[368:373],arar.prd$pred)*100
MAPEhw = mape(birth[368:373],HW.prd$mean)*100


MAPEs = c(fit1_mape = MAPE1,
          fit2_mape = MAPE2,
          fit3_mape = MAPE3,
          fit4_mape = MAPE4,
          fit5_mape = MAPE5,
          fitauto_mape = MAPEauto,
          arar.model = MAPEarar,
          hw.model = MAPEhw)
MAPEs
```

-   Fit5 (0,1,2)x(1,1,2)x12 has the lowest MAPE (Mean Absolute Percent Error) of 2.011.
