---
title: "STA6856 Time Series"
subtitle: "Assignment 2"
author: "Hector Gavilanes"
date: "9/28/2023"
format: html
self-contained: true
execute: 
  warning: false
  message: false
editor: visual
---

# Instructions:

**Submitting the same exact answer will lead to failing the assignment.**

## Problem A

Let $Z_t$ be a sequence of independent normal random variables, each with mean 0 and variance $\sigma^2$, and let $a$, $b$, and $c$ be constants. Which, if any, of the following processes are stationary? for each stationary process specify the mean, autocovariance, and autcorrelations functions.

1.  $X_t = a+ bZ_t + cZ_{t-2}$

2.  $X_t = a+ bZ_0$

```{r}
# Constants
a <- 10
b <- 5
c <- 2
sigma_squared <- 1

# Autocovariance for Process 1
autocov_process1 <- function(k) {
  if (k == 0 || k == 2) {
    return(c * sigma_squared)
  } else {
    return(0)
  }
}

# Display results for Process 1
cat("Problem 1:\n")
cat("Mean (μ):", a, "\n")
cat("Autocovariance (Cov) for lag 0:", autocov_process1(0), "\n")
cat("Autocovariance (Cov) for lag 2:", autocov_process1(2), "\n")

# Process 2 has zero autocovariance for all lags
cat("\nProblem 2:\n")
cat("Mean (μ):", a, "\n")
cat("Autocovariance (Cov) for any lag: 0\n")

```

## Problem B

Use the following time series:

-   `airpass{itsmr}`: Number of international airline passengers, 1949 to 1960.

-   `lake{itsmr}`: Level of Lake Huron, 1875 to 1972.

And perform the following:

1.  Plot and examine the time series and its ACF.

2.  Estimate the Trend and/or Season being considered using Moving Average Smoothing and Regression, if applicable

3.  Plot both trend estimates and the original series in one figure.

4.  Test the hypothesis whether or not the the residuals data are IID.

5.  State your conclusions whether or not further analysis should be pursued. Why?

6.  Use the difference operator to eliminate the Trend/Season components and redo steps 4 and 5.

```{r, warning=FALSE}
# Load packages
library(itsmr)
library(tidyverse)
library(tsibble)
library(feasts)
```

Load Time Series

```{r}

lake_data <- LakeHuron

```

```{r}
LakeHuron
lake_ts <- ts(LakeHuron, start = 1875, frequency = 1)
```

Plot the series

```{r}
# Plot the Air Passenger time series
plot(airpass, type="o", xlab = "Time", ylab="Passengers", 
     main="International Airline Passengers (1949 - 1960)")

# Plot ACF
acf(airpass, lag=100, col="2", xlab="Lag", ylab="Values", 
    main="ACF - Airline Passengers")

# Plot PACF
pacf(airpass, lag=100, col="2", xlab="Lag", ylab="Values", 
    main="ACF - Airline Passengers")
```

4.  IID test

```{r}
# IID test
adf.test(diff(log(airpass)), alternative="stationary")
```

5.  Use the Difference test

```{r}
# Plot the Air Passenger time series
plot(diff(airpass), type="o", xlab = "Time", ylab="Passengers", 
     main="Diff - International Airline Passengers (1949 - 1960)")

# Plot ACF
acf(diff(airpass), lag=100, col="2", xlab="Lag", ylab="Values", 
    main="Diff ACF - Airline Passengers")

# Plot PACF
pacf(diff(airpass), lag=100, col="2", xlab="Lag", ylab="Values", 
    main="Diff PACF - Airline Passengers")
```

```{r}
# Plot the Lake Huron time series
plot(lake_ts, main = "Lake Huron Water Level (1875-1972)", ylab = "Water Level")

# Plot autocorrelation function
acf(lake_ts, main = "ACF of Lake Huron Water Level")
```

```{r}
# transform data to tsibble format
water_level <- lake_ts %>% 
  as_tsibble(index = Year)

# Plot the time series
water_level %>% 
  autoplot() +
  labs(y="Water level",
       x="Time[1Y]",
       title = "Lake Huron Water Level (1875-1972)")

# Plot ACF
water_level %>% 
  ACF(water_level, lag_max = 100) %>% 
  autoplot()
```

Trend using Moving Average Smoothing

```{r}
ma_smooth <- stats::filter(lake_ts, sides=2, rep(1/10, 10))
plot(lake_ts, col = "black", main = "Lake Huron Water Level with Moving Average Smooth",
     ylab = "Water Level")
lines(ma_smooth, col = "magenta")
```

```{r}
water_level %>% 
  mutate(ma.trend3 = smooth.ma(value, 3),
         ma.trend5 = smooth.ma(value, 5))
pivot_longer(cols = -Year,
             names_to = "series",
             values_to = "Values") %>% 
  group_by(series) %>% 
  autoplot(values)
```

```{r}
# Estimate trend using moving average smoothing
water_level %>%
  mutate(ma.trend3 = smooth.ma(value, k = 3),
         ma.trend5 = smooth.ma(value, k = 5))

# Reshape data for plotting
water_level_long <- water_level %>%
  pivot_longer(cols = starts_with("ma.trend"), 
               names_to = "series", 
               values_to = "Values")

# Plot the trend estimates
water_level_long %>%
  ggplot(aes(x = Year, y = Values, color = series)) +
  geom_line() +
  labs(title = "Moving Average Smoothing for Trend Estimation",
       x = "Year", 
       y = "Trend Estimate") +
  theme_minimal()
```
